---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}
set.seed(42) # Setting seed for reproductibility
population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)
# View the first few rows
head(population)
```


### Task 3: Sample 100 rows of data from the dataset

```{r}
sample_100 <- population |> 
  sample_n(100)

sample_50 <- population |> 
  sample_n(50)

sample_10 <- population |> 
  sample_n(10)
```


### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample_100 |> summarize(mean = mean(age), sd = sd(age))

sample_50 |> summarize(mean = mean(age), sd = sd(age))

sample_10 |> summarize(mean = mean(age), sd = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population

```{r}
population |> summarize(mean = mean(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The results change based on sample sizes, as the largest sample size (100) depicts fairly accurate averages, but the mean and standard deviation get less accurate as the sample size goes down — with an outlier 41.2 mean for sample_10, compared to a more accurate 48.58 mean for sample_50 and 49.02 for sample_100. The respective means and standard deviations are similar, but get increasingly varied with smaller sample sizes as opposed to larger ones that are representative (and make up more) of the actual survey population. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
