---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?
_
Certain categories — such as accidents, accident with impound, auto stolen, theft and theft from auto — generagted higher per/month crime statistics than others over each of the 12 recorded months, revealing generally higher numbers for auto-related incidents in specific. From a reporting standpoint, I believe that these four categories are increasingly more important than numbers - as four of the highest crime statistic categories are auto/vehicle related, thus serving as an important trend that I believe should be further investigated. I don't believe this crime data makes the job of summarizing the data more difficult - if anything, I think this revealed trend related to automobile/vehicle crimes can be seen as a dominating outlier, thus making the job of separating, categorizing and summarizing the data less difficult.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
total_incidents <- pg_crime |> mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec)
pg_crime_averages <- total_incidents |> mutate(average_incidents = total_incidents/12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?
_
General accidents hold the highest average number of incidents per month at 481.3, followed by stolen automobiles, theft from auto and accident from impound — all four of which sit at averages of 320 incidents or more per month, relating to my earlier theme/trend of connected vehicle crime reports. Residential robbery, with several months without a reported incident, and homicides (5.83) have the lowest number of incidents per month — with both never reaching single-month double-digit incident totals, further illustrating the rarity of these two crimes within the area. This tells me that the overall pattern of crime in Prince George's County is heavility scaled and attributed to auto accidents and theft, which make up the dominating majority of reports. Robbery is a category that I have questions about, as I'd think average per/month incident reports would be higher specifically related to more frequent robbery crime, yet no robbery-related categories hold triple-digit per/month totals, with vehicle robberies even showing a decreasing crime rate throughout the latter half of the year.


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate(june_change = (average_incidents-Jun)/Jun * 100)
summer_crime_averages <- pg_crime_averages |> mutate(july_change = (average_incidents-Jul)/Jul * 100)
summer_crime_averages <- pg_crime_averages |> mutate(august_change = (average_incidents-Aug)/Aug * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

Overall, a majority of crime category averages increased throughout the summer months, rather than decreased. One pattern I noticed was the fact that a majority of violent crimes increased throughout these summer months of June, July August, with consistent double-digit totals of sex offenses, assault and assault shooting, among others, all resulting in increased incidents throughout these months. I believe this supports the theory that crime is more likely to occur in the summer months, as growing and elevated rates can be seen throughout this time of year - perhaps related to weather, and an increased population of individuals outside during these three months in specific. Overall, these increased patterns and trends generally support these theories - even though there are certain outliers.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Automobile and vehicle-related accidents saw elevated incident rates throughout Prince George's County during the entirety of this past year, accounting for the large majority of crimes as opposed to others.

General accidents, stolen automobiles, theft from auto and accident from impound make up the four highest crime totals throughout the county — all four of which sitting at averages of 320 reported incidents or more per month, serving as an important statistic to local citizens and individuals within the county - illustrating the fact that automobile-related incidents are on the rise as opposed to other, more violent crimes. In short, these monthly averages can be attributed to a sweeping majority of Prince George's County incidents, a primary crime issue within the county.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
