---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The salary data's split into four categories with more than 500 layered rows — describing name, status, salary and title, all of which have varying statistics/data. From a reporting standpoint, the fact that the entirety of the 14 highest-paid White House individuals are labeled as 'Detailees,' while the majority of high-paying salaries following it are general 'Employees' in assistant roles, most stands out to me — as the trend might leave a journalist/reporter wondering why and how these salaries/roles are shaped and interconnected. Yes, there are salary values that might cause problems for analysis — 11 individuals don't have any salary listed (shown as $0), possibly in volunteer roles. This could be problematic when analyzing the salary data as a whole, as there's no reason listed for specific salaries; furthermore, certain roles that depict a $0 salary (such as 'Senior Policy Adivsor') are duplicated later throughout the list with higher salaries, thus not matching this data. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(salary), median_salary = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>  summarize(mean_salary = mean(salary), median_salary = median(salary), min_salary = min(salary), max_salary = max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

Both the mean and median saalries tell me that the data is high and well-baalned — a high median salary (110,500) indicates a variety of both low and high-paying salaries balanced out throughout the data, while a six-figure mean salary (107,036) to match illustrates the fact that the majority of White House staff are paid/compensated generally well, not just a select few weighing the scale higher or lower — but that these salaries are reflected and indicative of a broader population of individuals. While the data is generally consistent, there are definitely outliers in the data — both minimum and maximum salary data provides drastic examples, with the former of which showing a 0 salary lower bound outlier and the maximum showing a 251,258 upper bound outlier.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

Based on the data, my best story idea revolves around the fact that some individuals were not compensated at all, while others with similar roles had increased salary numbers. Furthermore, I wonder about the White House pay disparity within specific individuals — what's the primary reason for such heavily-scaled and broadened salary ranges for similar job roles? The next step I might take to develop this idea is to track and categorize salaries by position/titles to check for similaries and correlation, then potentially reach out to related personnel to further track trends and reasons for pay disparity.
