---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 * 1000, violent_rate_2020 = violent2020/pop2020 * 1000, property_rate_2020 = property2020/pop2020 * 1000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

When comparing the cities with highest violent and property rates, I noticed a distinct similarity between both 2019 and 2020 — as Ocean City consitently ranked atop each respective list (first in 2019 property rate, 2020 property rate, 2020 violent rate; second in 2019 violent rate). The pattern can be related to generally similar cities ranking amongst the top of violent and property rates over 2019 and 2020 — with consistent appearances of Ocean City and Baltimore, followed by a differing variety of shuffling cities mostly following the two. I believe the story here is to explore Ocean City's dramatically high rates, and why Ocean City in specific has the highest violent and property rates over the span.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(violent_pct_change = (violent_rate_2020-violent_rate_2019)/violent_rate_2019 * 100, property_pct_change = (property_rate_2020-property_rate_2019)/property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The main pattern I noticed included generally decreasing percentage change rates, notably a dramatic 25% change in property percent and 12% in violent rate percent for Baltimore, while other outliers, such as Cambridge are more varied — with positive and negative correlations across violent and property categories, respectively. Furthermore, another pattern reveals increasingly positive change across violent percentage change rates from 2019 to 2020, as three of the top four percentage change rates for 2019 violent crime rates were negative values, while three of the top four rates for 2020 violent crime rates were positive — representing a gradual shift up between percentage change across the two years. A notable outlier included Edmonston's 895.32398130 violent percent change rate, 200 more than the next closest city, Poolesville (607.14698517). Both Edmonston and Poolesville ranked amongst the lower half of 2019 violent crime rates, while Edmonston in particular made a sizable jump to within the top-15 in 2020.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Ocean City consistently ranked amongst the top of property and violent crime rates for Maryland cities through 2019 and 2020, representing increasingly elevated rates compared to its fellow decreasing counterparts.

An 82% violent crime percent change rate and a 16% increased property crime percent change rate have gradually boosted Ocean City atop respective property and violent crime lists, as it ranked first in 2019 property rate (103.68), 2020 property crime rate (121.28) and 2020 violent crime rate (23.97), and second in 2019 violent crime rate (13.10). This data's important, and can serve as a representation of overall safety within the city, with increased crime over recent years suggesting Ocean City is more susceptible to crime and holds elevated, more crime incidents than other Maryland city counterparts.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: HW5 - Percentiles

```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile = ntile(violent_rate_2020, 100))
```

There are 11 total cities that sit at the 90th percentile or above, with 10 cities slotted above the 90th percentile.

```{r}
md_cities_percentiles |> summarize(IQR_violent_rate_2020 = IQR(violent_rate_2020))
```
Though the results may vary and create new story ideas, I don't believe it changes my best idea from the past assingment — to explore why Ocean City has dramatically high rates, and why Ocean City in specific has the highest violent and property rates over the span. The percentile data further supports this idea, as Ocean City ranks as the sole city in the 100th percentile — further showing its consistently high violent crime rates. While this is my best idea, another idea that came up after calculating percentile data could be centered around high violent crime rates within Prince George's County specifically — as the data reveals 5 of the top 15 (a third, 1/3) of the top percentiles came from PG County cities.
