---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
m_and_m <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}
set.seed(1234) # Setting seed for reproductibility
  sample_5 <- m_and_m |> 
    sample_n(5)
  
set.seed(1234) # Setting seed for reproductibility
  sample_10 <- m_and_m |> 
    sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.

```{r}
sample_5 |> summarize(mean = mean(red), sd = sd(red))

sample_10 |> summarize(mean = mean(red), sd = sd(red))
```


### Task 5: Calculate the mean and standard deviation of the original dataset, using the same color as Task 4

```{r}
m_and_m |> summarize(mean = mean(red), sd = sd(red))
```


### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Use a consistent binwidth.

```{r}
sample_5 |> 
  ggplot() +
  geom_histogram(aes(x = red), binwidth = 3) +
  geom_vline(aes(xintercept = mean(red)), color = "red", linetype = "dashed")

sample_10 |> 
  ggplot() +
  geom_histogram(aes(x = red), binwidth = 3) +
  geom_vline(aes(xintercept = mean(red)), color = "red", linetype = "dashed")

m_and_m |> 
  ggplot() +
  geom_histogram(aes(x = red), binwidth = 3) +
  geom_vline(aes(xintercept = mean(red)), color = "red", linetype = "dashed")
```


### Task 7: Answer some questions

Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?

While means, standard deviations and histograms are all helpful to me in understanding the accuracy of the samples, I'd say the histograms in specific best help me understand/analyze the sample data — as it provides a more visual representation of samples, showing clear outliers, skewed data and relative spread. For example, sample5 and sample10 both provide a limited baseline view of the data (largely stacked with clear outliers, lower and upper bounds without much of a consistent center), while the m&m population data shows a larger sample of data — that better identifies accuracy in relation to consistent findings within the dataset as a whole, rather than smaller, limited and less reliable/skewed samples.

What does this exercise tell you about the relationship between sample size and estimate accuracy?

This exercise tells me that in most cases, as the sample size increases, so estimate becomes more accurate as well. More variability and less reliable change is present with a smaller sample size — while with a larger sample size, estimates tend to be more accurate in relation to true population and exact data estimates/averages.

How does random sampling introduce variation into statistical estimates?

Random sampling introduces variation into statistical estimates because it provides a sense of randomness and unique spread — each chosen random sampling can provide different values, thus slightly changing statistical estimates while leading to different results, as opposed to staying exactly constant each time.

If you repeated this process, would the samples always be identical? Why or why not?

No, if I repeated this process, the samples wouldn't always be identical — as both data set categories sampled five and 10 respective random numbers, the spread of which could change if repeated. Each time a smaller sample of a population is chosen, the selected group can differ — while it would reduce in similar mean and standard deviation values, it likely would produce different numerical results each time, regardless of if it's from the exact same dataset or not. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
