---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "BEN GEFFNER"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

According to the correlation matrix and visualization summary of variables, certain variables that seem to be most correlated to voter turnout include "pct_white" (0.6932347) and "pct_unemployed" (-0.7393991) — though one is positive and the other negative, both boast strong correlation values overall. Furthermore, the "pct_bachelors_higher" (0.5398245) and "pct_white" (-0.6777275) variables seem to have the strongest correlation to population and turnout seems to be most correlated to "pct_white" (0.6932347); however, both are slotted below .70. On the other hand, variables relating to overall categories of income, pct_bachelors_higher and unemployed percentages generally seem to reflect greater correlation trends. A Bachelor's degree or higher, connected to educational background, has a strong positive 0.7520544 correlation to income, while the inverse predicts a mirrored result. Finally, five of the seven variables shown have a negative correlation to unemployment percentage, with voter turnout and pct_white being the strongest correlations. Specific relationships that show minimal/limited correlation between one another includes "bachelor's degree or higher %" to "median age,"income" to "median age" and "pct_white" to "income." Looking at the multiple scatterplots, certain variables that seem to have a linear relationship with the dependent variable includes unemployment percentage (negative) and pct_white (positive), while Bachelor's degree or higher percentage seems to have the lowest (positive, but extremely widespread with a plethora of outliers).

```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

Small spread minimum (-0.054193) and maximum (0.024262) values effectively suggest that the model fit avoids substantial errors in predictions for most cases, with an evenly distributed median value close to zero (0.000494) — these small residuals with little variation inspires confidence and reliability within the model. Furthermore, the adjusted R-square being so large (0.8023) suggests that these variables serve a high cause/correlation in relation to voter turnout — thus making it a useful model, in my opinion, to predict overall turnout relative to various factors. Furthermore, a small residual standard error (RSE) of 0.02174 indicates that the typical prediction by the model deviates from the actual turnout values by just 0.02174, or 2.174 percentage points — reaffirming a good fit with a small model spread with predictions that are increasingly close to actual data points. Though some coefficient variables effectively correlate to positive and increased voter turnout values, certain variables (such as pct_bachelors_higher, population and pct_unemployed) show very low or even negative relationship effects on voter turnout levels.

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

After creating three alternative models (with indpendent variables of population, pct_bachelors_higher and pct_white), I found the pct_white correlation (Model 4) to be the most significant relative to its voter turnout causation. While none of the three adjusted models came close to the full model's adjusted R-Square of 0.8023, pct_white was the next closest at 0.4569641 — significantly better than both population (0.1111636) and pct_bachelors_higher (0.05044749), both of which showed decreasingly low levels of correlation to voter turnout, thus making pct_white the strongest isolated factor relative to the dependent variable.

```{r}
# Model 2
model2 <- lm(turnout ~ population, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ pct_bachelors_higher, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

I could use this analysis to cover an election or voter participation story, from a broad sense, to recognize the impact of how various demographic/socioeconomic variables and factors relate to higher or lower overall voter turnout — as well as acknowledging patterns such as race, age and education level, among other factors, that play a part in limiting voter ability and turnout. When reporting on an voter participation story, finding a targeted storyline (such as economic challenges or education-related boundaries) could be key in comparing voter turnout from minority groups relative to overall election poll numbers — I'd search for both correlation/similarities and differences to narrow down the impact of factors and, in turn, simplify analysis for my story audience.
