---
title: "The Impact of Rest on NBA Team Performance"
NAME: "BEN GEFFNER"
---

## Perform ANOVA Test for '2024-25 Team Win Percentage Over Time' dataset (UPDATED as suggested without % symbols and adjusted to fit 'long-form' model on 5/3)

When performing my first ANOVA Test, my goal was to determine if there were real differences in the average rates of NBA rest impacts between different years, and if the results represented significance. Furthermore, I felt this team_over_time dataset would prove most significant for the ANOVA test format, as it holds a healthy sample size of over 300 total unique observations/win percentage rates, thus illustrating a diverse background of findings across recent history. The F-value is notably low at just .229, thus suggesting that the differences between these time periods don't outweigh the variation within each — and indicating little variance between group averages relative to each other. Furthermore, a high p-value much greater than the common .05 alpha tracking level indicates that the differences are not statistically significant, that there's no statistically significant difference in average win percentages across recent years. The absence of the three asterisks below the test fails to highlight that this is significant at the 0.001 level — bolstering these results with the finding that rest impacts haven't varied enough over time in connection to its respective effect on team win percentage — however, this doesn't necessarily indicate that no correlation between rest and win percent is present at all, but solely that recent yearly trends represent closely mirrored (and high relative to more rest) results with no dramatic rise or fall over time.

```{r anova-test}
# Run one-way ANOVA
anova_result_year <- aov(Win_Pct ~ Year, data = team_over_time)
summary(anova_result_year)
```

## Compare the differences between specific pairs

I thought the presence of a post-hoc test, in this case, a TukeyHSD test, would be useful to back up and confirm initial ANOVA year-by-year results — as well as to find which group means differ from each other, serving as more of a confirmation of the previous ANOVA Test results due to non-siginificant findings and a drastically high p-value result (0.99). Overall, it effectively confirmed what my ANOVA test already suggested to me, that there's no meaningful trend in rest impact by year specifically — but rather, it produced more constant and stable findings across all studied years from 2015 to 2025 (with the potential to be obstructed, of course, by outside factors such as a team's individual/unique success change over a 10-year period). Certain years, such as 2024-25 versus 2017-18, revealed above-average large mean differences but still noted high adjusted p-values that show no significant difference — in this case, the difference was 4.066667e-02, or 0.0407, but still held a p-adj of 0.997. 2017-18 consistently showed slightly lower win percentages relative to other years while 2024-25 mostly held higher, positive differences when comparing isolated findings to other years — but still not enough to show statistical significance for team over time results.

```{r post-hoc}
# Tukey's HSD test to see what differs
tukey_result_year <- TukeyHSD(anova_result_year)
print(tukey_result_year)
```

## Perform Linear Regression Model Analysis for '2024-25 Team Win Percentage Over Time' (UPDATED as suggested without % symbols on 5/3)

My final test for the team over time findings was a linear regression model analysis, testing whether win percentage could be predicted or influence by a specific year. In this case, I set win percentage as the dependent variable and the year as the independent variable — with hopes that if we know the year for a specific NBA season, the win percentage reliability within it would be reflected/revealed. The model reveals the intercept is 0.547667 with an array of reduced positive and negative yearly values surrounding it — further relating back to the non-significant trend with visible but unimportant differences seen relative to the yearly "over time" data. The second of three statistics mentioned in class shows the p-value as 0.9901 and the adjusted R-squared as -0.02375. Based on what I've learned this semester, it tells me that the nature of the relationship between yearly data and win percentages, based on minimum and maximum residual values, is less varied than my original hypothesis. In similar fashion to my previous tests, none of the years show a statistically meaningful change in win percentage, with an extremely low R-squared helping illustrate how the Year variable doesn't wholeheartedly predict win percentage consistently enough — however, this model does suggest, bolstered by similar intercept ranges, that NBA team win percentages have been largely stable year over year since 2015 with just minor positive and negative differences occurring sporadically.

```{r}
model <- lm(`Win_Pct` ~ `Year`, data = team_over_time)

summary(model)
```

## Perform ANOVA Test for '2024-25 Team Win Percentage by Rest Splits' dataset (UPDATED as suggested without % symbols on 5/3)

Furthermore, when performing my second ANOVA Test, my goal was to determine if there were real differences in the average rates of NBA rest impacts between different amounts/splits of rest times and days preceding games. I formulated NBA Team Win Percentage by Rest Splits data in the 2024-25 season, with the goal in using this specific dataset being to further confirm if NBA team win percentage relative to various types of rest advantages (in this case, variables included 'no rest,' 'one day's rest' and '2-3 day's rest') were significantly different from my personal expected frequencies noted in the original hypothesis. I initially was having trouble forming the tests (and receiving error messages), but found success after reformatting the data in my original Google Sheet to make it as 'long' a format as possible (more rows than columns) as suggested in recent feedback. Through the restructured data, I tested whether different chosen rest groups (rest_time) resulted/correlated in various team win percentages. The F-value is notably low at just 1.664, thus suggesting that, though there are positive and negative changes between groups, it generally indicates little differences between group averages relative to each other. Furthermore, a p-value of 0.195 — though much lower and verifiable than the year-by-year data — is not statistically significant at the .05 alpha tracking level, meaning that there's not enough evidence based on just this single test alone to conclude correlation.

```{r anova-test}
# Run one-way ANOVA
anova_result_splits <- aov(win_pct ~ rest_time, data = current_win_by_split)
summary(anova_result_splits)
```

## Compare the differences between specific pairs

Once again, I thought the presence of a post-hoc test, in this case, a TukeyHSD test, would be useful to back up and confirm initial ANOVA rest split results in showing which individual rest time group comparisons show significant differences in average team win percentage — as well as to find which group means differ from each other, serving as more of a confirmation of the previous ANOVA Test results. Overall, it effectively confirmed what my ANOVA test already suggested to me since p-values were greater than the 0.05 test level, that there's no statistically significant differences because of non-significant p-values alone. However, when I took a deeper look at the data of the whole, it's clear to see that though it doesn't appear officially significant from a statistical standpoint, there are smaller yet visible trends connected to (and supporting my hypothesis) the claim that more rest is associated with slightly higher win percentages — especially comparing the upper and lower bound categories of No Rest versus 2-3 Days Rest, in which there's a positive growth/uptick of nearly 8.5% between the two, respectively. Though it's not significant as a whole, each of the three baseline categories provide slightly noticable higher average win percentage rates with more rest time preceding games.

```{r post-hoc}
# Tukey's HSD test to see which specific pairs of college types differ
tukey_result_splits <- TukeyHSD(anova_result_splits)
print(tukey_result_splits)
```
## Visualization

This visualization below, reflective of in-class techniques and utilized formulas, helps illustrate and paint a picture of the gradual rise in NBA team win percentage relative to increased rest time (though not necessarily significant, still relative of a visible trend). Overall, it generally backs up my hypothesis in showing that the more rest and time off a team has prior to a game, the higher their respective win percentage will be. For example, in this case, average win percentages for NBA teams entering a game with no rest throughout the 2024-25 season was drastically below 0.50 (or 50%), while one and 2-3 rest days preceding games resulted in a spike/jump up to averages well above 0.50. In turn, this visualization helps show that in terms of rest splits specifically, varying amounts of rest preceding games certainly has some sort of impact in eventual win team percentage. My visualization, separated by these three rest split categories, accurately backed up and confirmed my original belief that an increased rest advantage between games generally results in somewhat of a higher, elevated mean percentage for teams overall (of course, with the potential to be heightened or reduced by outside factors such as strength of schedule, weighted success, etc.).

```{r} 
# Create chart (based on in-class formulas) that help show gradual rise in win % relative to increased rest time
ggplot(current_win_by_split, aes(x = factor(rest_time), y = win_pct)) +
  geom_boxplot(fill = "seagreen") +
  labs(title = "Win % by Rest Time",
       x = "Rest Time (days)",
       y = "Win Percentage")
```

## Perform Linear Regression Model Analysis for '2024-25 Team Win Percentage by Rest Splits' (UPDATED as suggested without % symbols on 5/3)

I completed a linear regression model analysis for the 2024-25 Rest Split findings as well, testing whether win percentage could be predicted or influence by different rest timetables leading upto games. In this case, I set win percentage as the dependent variable and rest time as the independent variable — with hopes that if we know the amount of rest for specific NBA games, the win percentage reliability within it would be reflected/revealed. The model reveals the intercept is 0.44800 with small yet positive rest time impact values surrounding it — while it still may be seen as non-significant, suggestive trends in win percentage increases is visible. The second of three statistics mentioned in class shows the p-value as 0.1953 and the adjusted R-squared as 0.0147 Based on what I've learned this semester, it tells me that the nature of the relationship between rest split data and win percentages, based on minimum and maximum residual values, is less varied than my original hypothesis — but still reflects a smaller-scale trend that supports it. Though not fully significant, rest time split data certainly shows some influence relative to unique NBA team win percentages — teams with 2–3 days rest had a +8.5% higher win percentage than those with no rest, a similar effect reflected in a +6.0% growth for the one day rest group.

```{r}
model <- lm(`win_pct` ~ `rest_time`, data = current_win_by_split)

summary(model)
```
## Model Refinement for '2024-25 Team Win Percentage by Rest Splits'

To further build and experiment with my regression model for 2024-25 NBA Rest Split data, I refined my model to add unique Team data along with win percentages and rest splits. The results showed a much improved new model — allowing for a much more precise element estimates to help explain team-by-team performances, respectively — bolstered by a 0.5696 adjusted R-square value, in direct contrast to a smaller and more limited adjusted R-Square of 0.0147 seen in my original model. After accounting for team differences and outliers now visible with the added variable in place, rest time solidifies itself as a meaningful indicator and predictor of win percentage rates, thus fully supporting my hypothesis — with this new model accurately strengthening results while simaltaneously accounting for team quality as another dominant factor that influences overall win rate. For example, even after taking into account rest time splits, objectively worse NBA teams such as Charlotte, Brooklyn, Portland, Utah and Washington show decreasingly negative intercept values — while it's now clear to see that rest time splits certainly represent a portion of influence, overall team strength is another predominant, if not leading, factor that must be acknowledged in my analysis. Teams with 2–3 days rest win significantly more often than those with no rest — with a general R-squared value representing how this model explains nearly 72% of the variation in team win percentage. This new model boasts increasingly high p-value (1.366e-07) and F-statistic (4.799) values which, in turn, better reflects my hypothesis while suggesting a very strong belief that this modified model explains variation in win percentage as a whole in 2024-25.

```{r}
# Build initial model with all predictors
initial_model <- lm(win_pct ~ rest_time + team, data = current_win_by_split)
summary(initial_model)

```
## Final Story Pitch

Through this story, I'd like to explore the tangible/noticeable connection between rest days and team performances, and how these two factors often go hand in hand in helping to paint the overall outcomes of specific games. Overall, NBA teams play better and win more with increased rest time before games (at least 2-3 days off) — as opposed to back-to-back games or games with just one day's rest. Furthermore, I'd argue that individualized and varying team strengths for various organizations around the NBA are nearly as valuable in predicting win percentage than increased rest time is.

Through research and data testing, as well as the eventual inclusion of 'team' variables via multiple regression analysis, I found that the impact of rest time on unique NBA win percentages can, in some ways, be seen as statistically meaningful. Boasting a p-value of 0.009, teams with 2-3 days rest prior to games showed a +8.5 percent boost in overall win percentage in 2024-25 in comparison to much lower no-rest numbers, while one-day rest rates showed a +6 percent improvement — both of which, backed by the above visualization, relate to the idea that there is a tangible and reflective correlation between increased rest days and elevated win percentages. Furthermore, one-way ANOVA and post-hoc tests comparing team win percentages by year (2015–16 to 2024–25) saw no significant differences across the 10-year data span, potentially suggesting trends in rest-related performance haven’t meaningfully changed over the last decade — however, though large differences cannot be verified by significance tests, there are certainly smaller individualized changes that have been evident across the past 10 years to further support potential story research in that team strengths for various organizations around the NBA are nearly as valuable in predicting win percentage. Certain upper bound teams — including Cleveland, Boston and Oklahoma City — consistently outperformed in 2024-25 specifically, while others such as Washington, Utah, New Orleans and Portland underperformed, regardless of rest — in turn, connecting back to levels of competition and overall team strengths. Additionally, isolating a team's rest advantage data over a 10-year span still, in some cases, provides a myriad of varying results despite constant/consistent rest data. The Brooklyn Nets, in specific, produced a 0.81 win rate 2020-21 and 0.70 in 2021-22 — a result of star-studded playoff rosters that performed well with Kevin Durant and others. Following those years, and subsequent trades and a roster overhaul that made Brooklyn one of the objectively weakest teams in basketball, the Nets have held a 0.41 win rate in 2023-24 and just 0.27 in 2024-25 — most certainly a product/result of strictly worsening team strength and outside factors, not more or less rest.

Potential problems and limitations include overall opponent and individualized game context — once again, relating to often overlooked outside factors that also join rest and team strength as variables that influence win percentage — as home and away game location and quality of opponent are currently not included in my model. Also, though my rest split data produced strong results, both the sample size and classification were lacking — limiting rest splits to just 2024-25 data might be a current road block in helping confirm trends, while expanding rest splits beyond just three categories (for example, total minutes played, travel distance during rest days to and from games, etc.) could assist in credibility and model sensitivity to account for hidden factors of influence. My reporting plan to deliver a story on how rest and team strength impacts NBA team performance would include spending time conducting background interviews with seasoned professionals in and around basketball (for example, coaches, players, sports scientists and nutritionists, team travel/operations directors, etc.), as well as better highlighting the reasons for certain teams performing relatively neutral with or without rest and presenting both a league-wide and nationwide baseline survey/poll to secure public thoughts on the topic. Furthermore, before continuing data exploration and writing the investigative story, I would explore the contexualized impact of load management, and how sitting star players for certain games factors into the equation of overall NBA team win percentages.

## FROM 4/26 — Brainstorming Final Project Tests

## Out of these brainstormed tests belows and shaped by instructor feedback, I chose to use an ANOVA test, linear and multiple regression model analyses, TukeyHSD test and a visualization.

Given various analyses experimented with in class, I believe tests relating to linear regression, ANOVA and chi-square would prove beneficial and be applicable for use when studying the accuracy of my hypothesis. First, I believe linear regression could be utilized well when connecting specific statistics represented in each of my datasets back to the overall relationship between the number of rest days versus overall season-long win percentage for each given team — with team execution/performance as the potential dependent variable and rest days/time off as the respective independent variable. Furthermore, an ANOVA test could work as well when studying the grouping of rest day categories (seen in the 'Rest Day Splits' dataset) and how more or less time/days off prior to games impacts team performance — this test could to see if mean/average team win percentage and overall margin of victory differs across these various categories/groups, helping reveal the exact significance of these differences while differentiating between potential real patterns with statistically significant results or random variation. Finally, a chi-square test could work in similar fashion — helping determine whether the frequency of a specific team's number of both losses and wins differs based on the amount of rest days preceding games, eventually testing how these two factors are either independent or interconnected, with the answer having the potential to confirm my original hypothesis in a more straightforward manner. This, in turn, could be very telling as to whether actual observed frequencies of NBA team win percentage relative to rest advantages are significantly different from my personal expected frequencies noted in the original hypothesis. A few questions I still have include whether it would be beneficial to run all of these above mentioned tests or just focus on a select few/one when comparing results to my original hypothesis, and if the results of these tests have the potential to be interconnected or if each specific finding should be isolated relative to each — as well as whether I'm on the right track in relation to accurately labeling specific variables that I plan to use for these tests.

## Exploring the effect of rest days on NBA team performances and the overall impact of in-season time off in between games.

Initial Hypothesis: 
## I believe that there is a noticeable connection/correlation between rest days and team performances, and that these two factors often go hand in hand in helping to paint the overall outcomes of specific games. Overall, NBA teams play better and win more with increased rest time before games (at least two days off) — as opposed to back-to-back games or games with just one day's rest.

### Load the tidyverse

```{r}
library(tidyverse)
```

```{r load-data 1 - 2024-25 With Rest Advantage (UPDATED as suggested without % symbols on 5/3)}
current_with_rest_ad <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTPN8ee-Ux4bjjcphB1UbT0zQuemOhamDLOfE-KumaO1fKN0enXk6SgQ8FyudVJpDLopO-egPb2Mua4/pub?output=csv")

# Preview structure
View(current_with_rest_ad)
```


```{r data-2 - 2024-25 With Rest Disadvantage (UPDATED as suggested without % symbols on 5/3)}
current_with_rest_dis <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vR73DN5g4n8xV_NDKYdp9zdojRjaFeC06O4H034KrpTTCxGHjpt0Ye8PSdHM3Zp_DvAWEWcBcZt7UK3/pub?output=csv")

# Preview structure
View(current_with_rest_dis)
```


```{r data-3 - 2024-25 Team Win Percentage by Rest Splits (UPDATED as suggested without % symbols on 5/3)}
current_win_by_split <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQMEP0XlbvoNFdNQpcsiTnHQdi6yh9HaM-1fZfreZ2gjbUJF5DAZn8f-3Ouf2M-bEayLqMDpyRxQtAP/pub?output=csv")

# Preview structure
View(current_win_by_split)
```
## While the latter 2 datasets highlight NBA team ineraction with rest days over a long period of time, the main focus of the first three is to isolate numbers based strictly from the 2024-25 season, as it provides the most up to date and current findings relative to my examination. I believe these initial three findings support my hypothesis that NBA teams playe better and win more with increased rest time before games — 21 of 30 NBA teams have winning records with a rest advantage, as illustrated in dataset 1, and more than half of all teams are over .500 in games following 2 days rest (time off) leading up to it, as seen in dataset 3. I believe column/dataset 3, specifically rest splits, will be most important for my analysis, as it allows me to compare specific team results (fluctuated and weighted by other factors, of course, like team overall win rate and strength) while isolating different rest split classes, moving from no rest to 2-3 days rest. 

One question I still have on this is how significant is overall rest versus/compared to just overall team quality, and how road versus home game stretches impact rest results? I might be missing specific other variables that are much tougher to account for, such as travel distance during road rest time, altitude, back-to-backs leading into rest days, etc.

```{r data-4 - Team Win Percentage With Rest Advantage Over Time}
team_over_time <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSiVEJFX54kzR_PGFXM8ewhYyru-cAZ180ilNIxDBZNr9e6SU6L6wzzG22J74qDv-wAQMuMdhWtkGV9/pub?output=csv")

# Preview structure
View(team_over_time)
```


```{r data-5 - Rest Advantage Comparison By Time Period}
total_by_time_period <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRK56dBfbzYVb8vYiONb9Sf4H1f4jajM393iHDzw_SXNhgGvXyA_VbU4-bmmlzgU6sWsBe13t49-cnf/pub?output=csv")

# Preview structure
View(total_by_time_period)
```

## These last two datasets help paint the picture and positive impact of rest days on NBA team performance on a much more widespread, broader angle over a large period of time. This NBA rest history data, tracked back to 2003, will definitely be helpful in relating and confirming my beliefs/hypothesis beyond just the 2024-25 season.

Some questions that these last two datasets might help solve include how closely long-term NBA post-rest game performance trends correlates with the current 2024-25 season trends, as well as how these historical numbers can assist in prediciting future patterns and rest impacts.
