---
title: "The Impact of Rest on NBA Team Performance"
NAME: "BEN GEFFNER"
---

Given various analyses experimented with in class, I believe tests relating to linear regression, ANOVA and chi-square would prove beneficial and be applicable for use when studying the accuracy of my hypothesis. First, I believe linear regression could be utilized well when connecting specific statistics represented in each of my datasets back to the overall relationship between the number of rest days versus overall season-long win percentage for each given team — with team execution/performance as the potential dependent variable and rest days/time off as the respective independent variable. Furthermore, an ANOVA test could work as well when studying the grouping of rest day categories (seen in the 'Rest Day Splits' dataset) and how more or less time/days off prior to games impacts team performance — this test could to see if mean/average team win percentage and overall margin of victory differs across these various categories/groups, helping reveal the exact significance of these differences while differentiating between potential real patterns with statistically significant results or random variation. Finally, a chi-square test could work in similar fashion — helping determine whether the frequency of a specific team's number of both losses and wins differs based on the amount of rest days preceding games, eventually testing how these two factors are either independent or interconnected, with the answer having the potential to confirm my original hypothesis in a more straightforward manner. This, in turn, could be very telling as to whether actual observed frequencies of NBA team win percentage relative to rest advantages are significantly different from my personal expected frequencies noted in the original hypothesis. A few questions I still have include whether it would be beneficial to run all of these above mentioned tests or just focus on a select few/one when comparing results to my original hypothesis, and if the results of these tests have the potential to be interconnected or if each specific finding should be isolated relative to each — as well as whether I'm on the right track in relation to accurately labeling specific variables that I plan to use for these tests.

## Exploring the effect of rest days on NBA team performances and the overall impact of in-season time off in between games.

Initial Hypothesis: 
## I believe that there is a noticeable connection/correlation between rest days and team performances, and that these two factors often go hand in hand in helping to paint the overall outcomes of specific games. Overall, NBA teams play better and win more with increased rest time before games (at least two days off) — as opposed to back-to-back games or games with just one day's rest.

```{r load-data 1 - 2024-25 With Rest Advantage}
current_with_rest_ad <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQMoS7M3DyC3zMLcPNIKftbe4PGtgpYH6GnjSnP2Z_pq4BdGkLC_bCbYrxlVF7wKTfAuAh-Zg_N_4Ry/pub?output=csv")

# Preview structure
View(current_with_rest_ad)
```


```{r data-2 - 2024-25 With Rest Disadvantage}
current_with_rest_dis <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vS0bzQtZKkNbFDJm8UfbYv7lrDfkl1LMrC--mLiPX4XJa2OOWFWJp5VzIgcB6ohIvZ0b2uwNgtpB-AA/pub?output=csv")

# Preview structure
View(current_with_rest_dis)
```


```{r data-3 - 2024-25 Team Win Percentage by Rest Splits}
current_win_by_split <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRhvy8mewisID7_XudWPpibJKdJ25NodqeIew6uUTc8fFa75zeDZix39kzIhLRzZ5jSnL7ocQaWxA17/pub?output=csv")

# Preview structure
View(current_win_by_split)
```
## While the latter 2 datasets highlight NBA team ineraction with rest days over a long period of time, the main focus of the first three is to isolate numbers based strictly from the 2024-25 season, as it provides the most up to date and current findings relative to my examination. I believe these initial three findings support my hypothesis that NBA teams playe better and win more with increased rest time before games — 21 of 30 NBA teams have winning records with a rest advantage, as illustrated in dataset 1, and more than half of all teams are over .500 in games following 2 days rest (time off) leading up to it, as seen in dataset 3. I believe column/dataset 3, specifically rest splits, will be most important for my analysis, as it allows me to compare specific team results (fluctuated and weighted by other factors, of course, like team overall win rate and strength) while isolating different rest split classes, moving from no rest to 2-3 days rest. 

One question I still have on this is how significant is overall rest versus/compared to just overall team quality, and how road versus home game stretches impact rest results? I might be missing specific other variables that are much tougher to account for, such as travel distance during road rest time, altitude, back-to-backs leading into rest days, etc.

```{r data-4 - Team Win Percentage With Rest Advantage Over Time}
team_over_time <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSiVEJFX54kzR_PGFXM8ewhYyru-cAZ180ilNIxDBZNr9e6SU6L6wzzG22J74qDv-wAQMuMdhWtkGV9/pub?output=csv")

# Preview structure
View(team_over_time)
```


```{r data-5 - Rest Advantage Comparison By Time Period}
total_by_time_period <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRK56dBfbzYVb8vYiONb9Sf4H1f4jajM393iHDzw_SXNhgGvXyA_VbU4-bmmlzgU6sWsBe13t49-cnf/pub?output=csv")

# Preview structure
View(total_by_time_period)
```

## These last two datasets help paint the picture and positive impact of rest days on NBA team performance on a much more widespread, broader angle over a large period of time. This NBA rest history data, tracked back to 2003, will definitely be helpful in relating and confirming my beliefs/hypothesis beyond just the 2024-25 season.

Some questions that these last two datasets might help solve include how closely long-term NBA post-rest game performance trends correlates with the current 2024-25 season trends, as well as how these historical numbers can assist in prediciting future patterns and rest impacts.
